Detecting ditches using \hl{supervised} learning on high-resolution \hl{digital elevation models}

### 2021-01-16 LaTeX ESWA/article.tex
Basically completes the introduction.

\begin{abstract}
{\footnotesize
\hl{Drained wetlands can constitute a large source of greenhouse gas emissions, but the drainage networks in these wetlands are largely unmapped, and better maps are needed to predict the climate consequences and aid in forest management decisions.} We develop a method for detecting ditches in high resolution digital elevation models derived from LiDAR scans. \hl{Thresholding methods using digital terrain indices can be used to detect ditches,} however, a single threshold generally does not capture the variability in the landscape and generates many false positives and negatives. We hypothesise that, by combining the digital terrain indices using \hl{supervised} learning, we can improve ditch detection at a landscape-scale. In addition to \hl{digital} terrain indices, additional features are generated by transforming the data to include neighbouring cells \hl{for better ditch predictions}. \hl{A Random Forests classifier} is used to locate the ditches, and \hl{its probability output} is processed to remove noise, and binarised to produce the final ditch prediction. The confidence interval for the Cohen's Kappa index ranges \textbf{[0.655 , 0.781]} between the evaluation plots with a confidence level of \textbf{95\%}. The study demonstrates that combining information from  a suite of digital terrain indices using machine learning provides an effective technique for automatic ditch detection at a landscape-scale\hl{, aiding in both practical forest management and in combatting climate change}.}
\end{abstract}

@@

Several previous studies have located ditches from Light Detection and Ranging (LiDAR) data in smaller agricultural areas \citep{roelens, bailly} or smaller forest areas (smaller than 2 $km^2$) \citep{rapinel, kiss}, predominantly characterised by artificial drainage. In this study, however, we conduct a landscape-scale analysis in a mesoscale catchment. The study  catchment  is 68 $km^2$, representing diverse landscape features and geomorphological characteristics. Additionally, the forest landscape introduces many challenges that are not present when detecting ditches along roads or in agricultural areas. The canopy cover in forest areas prevents the use of satellite imagery or aerial photos, as well as lowers the quality of the digital elevation models (DEMs), as fewer LiDAR points reach the ground, instead frequently catching in trees, bushes, or grass. Forest ditches are also generally not maintained to the same extent as agricultural or road ditches, often being overgrown or covered by windthrows or windsnaps. Using raw LiDAR point clouds to detect ditches, as \citet{roelens} and \citet{bailly} did in their studies can thus be problematic in a forest landscape, where a DEM constructed to map the points on the ground should be more suitable. \citet{cazorzi} used a DEM with image processing techniques aimed at detecting the elongated ditch shapes and excluding large landscape features. This approach was suitable for their open agricultural areas, but is difficult to generalise to a diverse forest landscape. Previous studies have found that terrain indices such as Sky View Factor \citep{zaksek}, Impoundment Index \citep{whiteboxtools} and High Pass Median Filter \citep{whiteboxtools} can detect ditches in forest landscapes in some cases \citep{uppsala}. However, different false positives and negatives can be observed when examining each terrain index manually. These inaccuracies can stem either from the gaps in the LiDAR scans, or from a poorly set threshold for the terrain index parameters. Thus, it would be useful to combine these LiDAR derived terrain indices to prune the weaknesses, and amplify the strengths of each index and produce highly accurate detection of ditch networks.

In this study, we introduce a new method for the automated mapping of ditches by integrating multiple high resolution terrain indices using a supervised learning approach. Several features and post-processing steps, specifically tailored for ditch detection are also developed and employed. These features aim to both highlight elongated shapes similarly to \citet{cazorzi}, as well as include a map of the neighbouring area as \citet{roelens} did in their study. We hypothesise that by combining the information from all the digital terrain indices using a machine learner (Random Forests), we can improve the detection of the ditches, compared to if a single terrain index is used. To test this, we conduct a landscape-scale ditch detection analysis at the Krycklan Catchment in Sweden \citep{krycklancatchment}, where a high precision manual mapping has been performed previously. This manual mapping guided the model training and ground-truthing in our analysis and provided a test-bench for our methodology.

@@

\subsubsection{Training and validation datasets}
\label{trainingvalidationdatasets}
To develop and evaluate our ditch detector, the raster and ditch label data of Krycklan were manually divided into 21 subsections. \hl{This was necessary because our method relies on the pixels surrounding a pixel, meaning it is not possible to select random pixels for training and testing.} Each subsection represents an area of roughly 196 hectare. From this division, 11 of the subsections were put aside as hold-out data, only for use in the experiment to evaluate the performance of the predictions. The remaining 10 subsections were used solely before the experiment to develop the ditch detector to test different ways of preparing the data. This allowed the ditch detector to be evaluated on  independent data and  strengthened the validity of the experiment. \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows which subsections were used for development and evaluation respectively. \hl{The 11 hold-out subsections were used as folds in an 11-fold cross validation in the final experiment.}

@@

\begin{table} [!htb]
        {\begin{tabular}{r|ll}
          Pos & Variable\textsuperscript{a} & Importance (\%) \\
          \hline
          1.  & Impoundment mean 3                                  & 7.33\\
          2.  & HPMF mean 4                                         & 6.23\\
          3.  & Impoundment mean 4                                  & 5.57\\
          4.  & Impoundment mean 2                                  & 5.41\\
          5.  & HPMF mean 3                                         & 4.33\\
          6.  & Impoundment median 4                                & 3.84\\
          7.  & HPMF Gabor - streams removed                        & 3.54\\
          8.  & HPMF median 4                                       & 3.40\\
          9.  & Impoundment median 2                                & 3.01\\
          10. & Sky View Factor Gabor - streams removed             & 2.84\\
          11. & HPMF mean 6                                         & 2.73\\
          12. & Impoundment median 6                                & 2.69\\
          13. & Impoundment standard deviation 4                    & 2.52\\
          14. & Sky View Factor Gabor                               & 2.44\\
          15. & Impoundment ditch amplification                     & 2.43\\
          16. & Impoundment mean 6                                  & 2.27\\
          17. & Impoundment ditch amplification - streams removed   & 2.24\\
          18. & HPMF min 2                                          & 2.23\\
          19. & Slope standard deviation 6                          & 2.01\\
          20. & Sky View Factor non-ditch amplification             & 1.88\\
          \hline
        \end{tabular}}
        \caption{The top 20 features (out of 40) by importance when the models make a prediction. \newline \textsuperscript{a} The number next to some of the variables indicates the circular radius used to select what neighbouring pixels to use in the statistical aggregation method. The radii represents pixels with a 0.5 m resolution.}
    \label{featureimportancetable}
\end{table}
\clearpage

>>>>

\begin{abstract}
{\footnotesize
\hl{Drained wetlands can constitute a large source of greenhouse gas emissions, but the drainage networks in these wetlands are largely unmapped, and better maps are needed to aid in forest management decisions, and to predict the climate consequences.} We develop a method for detecting ditches in high resolution digital elevation models derived from LiDAR scans. \hl{Thresholding methods using digital terrain indices can be used to detect ditches,} however, a single threshold generally does not capture the variability in the landscape, and generates many false positives and negatives. We hypothesise that, by combining the digital terrain indices using \hl{supervised} learning, we can improve ditch detection at a landscape-scale. In addition to digital terrain indices, additional features are generated by transforming the data to include neighbouring cells \hl{for better ditch predictions}. \hl{A Random Forests classifier} is used to locate the ditches, and \hl{its probability output} is processed to remove noise, and binarised to produce the final ditch prediction. The confidence interval for the Cohen's Kappa index ranges \textbf{[0.655 , 0.781]} between the evaluation plots with a confidence level of \textbf{95\%}. The study demonstrates that combining information from  a suite of digital terrain indices using machine learning provides an effective technique for automatic ditch detection at a landscape-scale\hl{, aiding in both practical forest management and in combatting climate change}.}
\end{abstract}

@@

Several previous studies have located ditches from remote sensing Light Detection and Ranging (LiDAR) data in smaller agricultural areas \citep{roelens, bailly} or smaller forest areas (smaller than 2 $km^2$) \citep{rapinel, kiss}, predominantly characterised by artificial drainage. In this study, however, we conduct a landscape-scale analysis in a mesoscale catchment. The study  catchment  is 68 $km^2$, representing diverse landscape features and geomorphological characteristics. Additionally, the forest landscape introduces many challenges that are not present when detecting ditches along roads or in agricultural areas. The canopy cover in forest areas prevents the use of satellite imagery or aerial photos, as well as lowers the quality of the digital elevation models (DEMs), as fewer LiDAR points reach the ground, instead frequently catching in trees, bushes, or grass. Forest ditches are also generally not maintained to the same extent as agricultural or road ditches, often being overgrown or covered by windthrows or windsnaps. Using raw LiDAR point clouds to detect ditches, as \citet{roelens} and \citet{bailly} did in their studies can thus be problematic in a forest landscape, where a DEM constructed to map the points on the ground should be more suitable. \citet{cazorzi} used a DEM with image processing techniques aimed at detecting the elongated ditch shapes and excluding large landscape features. This approach worked well for their open agricultural areas, but is difficult to generalise to a diverse forest landscape. Instead, we will focus on the topographical incision of the ditches in the DEM, which will be highlighted with a set of digital terrain indices. Previous studies have found that terrain indices such as Sky View Factor \citep{zaksek}, Impoundment Index \citep{whiteboxtools} and High Pass Median Filter \citep{whiteboxtools} can detect ditches in forest landscapes in some cases \citep{uppsala}. However, different false positives and negatives can be observed when examining each terrain index manually. These inaccuracies can stem either from the gaps in the LiDAR scans, or from a poorly set threshold for the terrain index parameters. Thus, it would be useful to combine these LiDAR derived terrain indices to prune the weaknesses, and amplify the strengths of each index and produce highly accurate detection of ditch networks.

In this study, we introduce a new method for the automated mapping of ditches by integrating multiple high resolution terrain indices using a supervised learning approach. Several features and post-processing steps, specifically tailored for ditch detection are also developed and employed. These features aim to both highlight elongated shapes similarly to \citet{cazorzi}, as well as include a map of the neighbouring area as \citet{roelens} did in their study. We hypothesise that by combining the information from all the digital terrain indices using a machine learner (Random Forests), we can improve the detection of the ditches, compared to if a single terrain index is used. To test this, we conduct a landscape-scale ditch detection analysis at the Krycklan Catchment in Sweden \citep{krycklancatchment}, where a high precision manual ditch mapping has been performed. This manual mapping guided the model training and ground-truthing in our analysis and provided a test-bench for our methodology.

@@

\subsubsection{Training and validation datasets}
\label{trainingvalidationdatasets}
To develop and evaluate our ditch detector, the raster and ditch label data of Krycklan were manually divided into 21 subsections. \hl{This was necessary because our method relies on the area surrounding a pixel being examined, meaning it is not possible to select random pixels for training and testing.} Each subsection represents an area of roughly 196 hectare. From this division, 11 of the subsections were put aside as hold-out data, only for use in the experiment to evaluate the performance of the predictions. The remaining 10 subsections were used solely before the experiment to develop the ditch detector to test different ways of preparing the data. This allowed the ditch detector to be evaluated on  independent data and  strengthened the validity of the experiment. \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows which subsections were used for development and evaluation respectively. \hl{The 11 hold-out subsections were used as folds in an 11-fold cross validation in the final experiment.}

@@

\begin{table} [!htb]
\centering
    {\begin{tabular}{r|ll}
      Pos & Variable\textsuperscript{a} & Importance (\%) \\
      \hline
      1.  & Impoundment mean 3                                  & 7.33\\
      2.  & HPMF mean 4                                         & 6.23\\
      3.  & Impoundment mean 4                                  & 5.57\\
      4.  & Impoundment mean 2                                  & 5.41\\
      5.  & HPMF mean 3                                         & 4.33\\
      6.  & Impoundment median 4                                & 3.84\\
      7.  & HPMF Gabor - streams removed                        & 3.54\\
      8.  & HPMF median 4                                       & 3.40\\
      9.  & Impoundment median 2                                & 3.01\\
      10. & Sky View Factor Gabor - streams removed             & 2.84\\
      11. & HPMF mean 6                                         & 2.73\\
      12. & Impoundment median 6                                & 2.69\\
      13. & Impoundment standard deviation 4                    & 2.52\\
      14. & Sky View Factor Gabor                               & 2.44\\
      15. & Impoundment ditch amplification                     & 2.43\\
      16. & Impoundment mean 6                                  & 2.27\\
      17. & Impoundment ditch amplification - streams removed   & 2.24\\
      18. & HPMF min 2                                          & 2.23\\
      19. & Slope standard deviation 6                          & 2.01\\
      20. & Sky View Factor non-ditch amplification             & 1.88\\
      \hline
    \end{tabular}}
    \caption{The top 20 features (out of 40) by importance when the models make a prediction. \newline \textsuperscript{a} The number next to some of the variables indicates the circular radius used to select what neighbouring pixels to use in the statistical aggregation method. The radii represents pixels with a 0.5 m resolution.}
    \label{featureimportancetable}
\end{table}
\clearpage
